Prediction of functional markers of mass cytometry data via deep learning

### 2018-11-14 manuscript/arxiv/nn-cytof.tex
finished paper

%%%%%%%%%%%%%%%%%%%% author.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% sample root file for your "contribution" to a contributed volume
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%% Springer %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% RECOMMENDED %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[graybox]{svmult}

% choose options for [] as required from the list
% in the Reference Guide

\usepackage{mathptmx}       % selects Times Roman as basic font
\usepackage{helvet}         % selects Helvetica as sans-serif font
\usepackage{courier}        % selects Courier as typewriter font
\usepackage{type1cm}        % activate if the above 3 fonts are
                            % not available on your system
%
\usepackage{makeidx}         % allows index generation
\usepackage{graphicx}        % standard LaTeX graphics tool
                             % when including figure files
\usepackage{multicol}        % used for the two-column index
\usepackage[bottom]{footmisc}% places footnotes at page bottom

% see the list of further useful packages
% in the Reference Guide

\makeindex             % used for the subject index
                       % please use the style svind.ist with
                       % your makeindex program

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% track changes
\usepackage[normalem]{ulem}
\usepackage[dvipsnames]{xcolor}
\newcommand{\claudia}[2]{\sout{#1}\textcolor{magenta}{\textbf{#2}}}
\newcommand{\missing}[1]{\textcolor{red}{\textbf{#1}}}

@@

\title*{Prediction of functional markers of mass cytometry data via
  deep learning}
% Use \titlerunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
%\thanks{Both authors contributed equally to this manuscript}
\author{Claudia Sol\'is-Lemus, Xin Ma, Maxwell Hostetter II, Suprateek
  Kundu, Peng Qiu, Daniel Pimentel-Alarc\'on}
% Use \authorrunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\institute{Claudia Sol\'is-Lemus \at Emory University, Atlanta, GA, \email{csolisl@emory.edu}
\and Xin Ma \at Emory University, Atlanta, GA,
\email{xin.ma@emory.edu}
\and Maxwell Hostetter II \at Georgia State University, Atlanta, GA,
\email{mhostetter1@student.gsu.edu}
\and Suprateek Kundu \at Emory University, Atlanta, GA,
\email{suprateek.kundu@emory.edu}
\and Peng Qiu \at Georgia Institute of Technology, Atlanta, GA,
\email{peng.qiu@bme.gatech.edu}
\and Daniel Pimentel-Alarc\'on \at Georgia State University, Atlanta, GA,
\email{pimentel@gsu.edu}}
%
% Use the package "url.sty" to avoid
% problems with special characters
% used in your e-mail or web address
%
\maketitle

\abstract{add here}

@@

Despite the technological advances in acquiring an increasing number
of parameters per single cell, approaches for analyzing such complex
data lag behind. The existing approaches are often subjective and
labor-intensive. For example, the widely used gating approach
identifies cell types by user-defined sequences of nested 2-D
plots. There have been efforts to develop clustering algorithms (e.g.,
flowMeans\missing{(reference)}, flowSOM\missing{(reference)},
X-shift\missing{(reference)}) and dimension reduction algorithms
(e.g., SPADE\missing{(reference)}, tSNE\missing{(reference)},
Scaffold\missing{(reference)}). However, there is still huge space
for developing new methods to ask new questions in this field.
% ---

@@

Furthermore, CyTOF data is perfectly suited for deep learning methods.
On one side, identify markers define a cell type (e.g., B cell, T
cell, monocytes, MSC), and on the other side, expressions of
functional markers identify the cell's activity (e.g., quiescent,
secreting cytokines, proliferating, apoptosis). Since CyTOF technology
allows for the simulteanous measurement of a large number of protein
markers, most CyTOF studies measure both identity markers and
functional markers, providing data for supervised learning tools, like
neural networks.  In addition, each CyTOF run typically collects data
on $10^6$ cells, creating an ideal large dataset in which the number
of samples (cells) is orders of magnitued larger than the number of
variables (markers). Deep learning methods are particularly suited for
this type of big data.

@@

@@

The dataset has been previously published in
\cite{Bendall2011,Qiu2011}. It contains measurements for 13 identity
(surface) markers, and 18 functional markers, which were performed on
5 healthy untreated donor bone marrow samples. The data roughly
contains 250,000 cells per sample (1,223,228 cells in
total). Thus, the data can be expressed in a $1223228 \times 31$
matrix.

The data was transformed with inverse hyperbolic sine function
($arcsinh$ with co-factor of 5), which is the standard transformation
for CyTOF data\missing{(reference)}.

\missing{(Peng: what was done to the data prior to sharing with us?)}

@@

We will compare the performance different methods (explained next
section) to predict the functional markers with the surface markers.
The data is highly complex and correlated, violating some of the
fundamental assumptions of standard statistical approaches (like
regression). For example, the data is highly skewed and the pattern
between response and predictor is not linear (see figure
\ref{exploratory}).

\begin{figure}
\includegraphics[scale=0.3]{figures/surface-hist.pdf}
\includegraphics[scale=0.3]{figures/functional-hist.pdf}
\includegraphics[scale=0.3]{figures/scatterplot.pdf}
\caption{Exploratory plots of surface and functional markers. The
  histograms show a biased pattern, and the scatterplot shows non
  linearity, both violations of crucial assumptions in standard
  regression models.}
\label{exploratory}
\end{figure}

@@

The complete data consisted of 1,223,228 cells in 18 functional
markers (responses) and 15 surface markers (predictors), which we
divided as follows: 250,000 rows as training set, 250,000 rows as
validation set, and 223,228 rows as testing set.

@@

The vector MSE is defined as
\begin{equation}
MSE_{vec} = \frac{1}{n} \sum_{i=1}^n ||\hat{Y}_i - Y_i ||^2_2
\label{vectorMSEformula}
\end{equation}
where $\hat{Y}_i \in \mathbf{R}^{18}$ is the predicted vector of
responses for individual $i$, and $Y_i \in \mathbf{R}^{18}$ is the
observed vector of responses for individual $i$.

The individual MSE for predictor $k$ is defined  as
\begin{equation}
MSE_{(k)} = \frac{1}{n} \sum_{i=1}^n (\hat{Y}_{k,i} - Y_{k,i} )^2
\label{MSEformula}
\end{equation}
where $\hat{Y}_{k,i} \in \mathbf{R}$ is the $k^{th}$ predicted
response ($k=1,\cdots,18$) for individual $i$, and $Y_{k,i} \in
\mathbf{R}$ is the $k^{th}$ observed response for individual $i$.

% https://github.com/pluskid/Mocha.jl/blob/master/docs/user-guide/solver.rst
\vspace{0.25cm}
\noindent \textbf{Neural network model:} We tested different network
architectures, activation functions, regularization coefficients,
solver methods, momentum policies, and learning rates. The best
network has four layers (see figure \ref{nn}) with 90, 90, 45 and 45
nodes. The network uses hyperbolic tangent as activation function,
regularization coefficient of 0.0001, momentum policy at 0.8, learning
rates of 0.01, 0.0001, 0.75. We used Adam solver\cite{Kingma2014}, an
algorithm for first-order gradient-based optimization of stochastic
objective functions, based on adaptive estimates of lower-order
moments, instead of Stochastic Gradient Descent (SGD) as the former
showed increased accuracy. All networks were trained using the julia
package Mocha\missing{(reference)}.

\begin{figure}
\centering
\includegraphics[scale=0.4]{figures/neuralNet.eps}
\caption{Neural network for predicting functional markers (18
  responses) from surface markers (15 predictors) with 4 hidden
  layers with 90, 90, 45 and 45 nodes each.}
\label{nn}
\end{figure}

\vspace{0.25cm}
\noindent \textbf{Linear regression (unpenalized/penalized):} We fit
standard linear regression, as well as the penalized version with
LASSO penalty under different penalization parameters. We used the
\textit{ScikitLearn}\missing{(reference)} julia wrapper, with default
settings. We noted that the penalized version performed worse than the
unpenalized version for all the predictors (regardless of penalty
parameter), so we only present results below for the unpenalized
linear regression model.

\vspace{0.25cm}
\noindent \textbf{Decision tree and random trees:} We fit one decision
regression tree with \textit{ScikitLearn} julia wrapper, with default
settings. We compared the performance of the ``mse'' criterion and the
Friedman's improvement score, deciding on the former (``mse'') which
is the default setting. We did not constraint the maximum depth of the
tree, and set as 2 the minimum number of samples required to split an
internal node. In addition, we did not constraint the maximum number
of features to consider when looking for the best split. Later, we fit
20 trees into a random forest method. We could explore more than 20
trees due to computational time constraints.

@@

\begin{figure}
\centering
\includegraphics[scale=0.35]{figures/vector-mse.pdf}
\includegraphics[scale=0.35]{figures/time.pdf}
\caption{Left: Vector MSE (equation \ref{vectorMSEformula}) for all
  four methods sorted from most accurate (neural network) to least
  accurate (decision tree). Right: Running time (in seconds) for the
  training and validation sets (sample 750,000 rows) for all four
  methods, sorted from fastest (linear regression) to slowest (random
  forest).}
\label{vectorMSE}
\end{figure}

Figure \ref{MSE} shows the individual MSE (equation \ref{MSEformula})
per predictor (18 predictors in x-axis) for each of the four
methods. Again, the prediction accuracy of neural network is better
than the other three methods for all the 18 predictors.

\begin{figure}
\centering
\includegraphics[scale=0.45]{figures/mse.pdf}
\caption{Individual MSE (equation \ref{MSEformula}) for each 18
  predictors. The neural network outperforms all other methods across
  all predictors.  Lines are drawn simply for visual effect.}
\label{MSE}
\end{figure}

Finally, we present a figure of the predicted response vs the observed
response, for all the four methods and 18 functional markers (figure
\ref{predObs}).

\begin{figure}
\centering
%\includegraphics[scale=0.4]{figures/neuralNet.eps}
\caption{Predicted vs observed responses across all functional markers
  and all four methods. The closer the slope to 1, the
  better.\missing{(missing this plot)}.}
\label{predObs}
\end{figure}

@@

The improved prediction accuracy of neural networks can be explained
by their flexibility to account for non-linearity or skewness.  Unlike
regression models, neural networks do not have linearity or normality
assumptions, and they take advatange of the correlation structure
among responses by fitting a network for the whole response vector.

As mentioned already, CyTOF data is perfectly suited for deep learning
methods given the simulteanous measurement of a large number of
protein markers, including both both identity markers and functional
markers. Both measurements allow for the implementation of highly
accurate supervised methods, like neural networks. In addition, the
structure of CyTOF data is ideal for deep learning: number of samples
orders of magnited greater than the number of variables.

@@

>>>>

\documentclass[conference]{IEEEtran}

%========================================================================= PACKAGES
\usepackage{amsmath,amssymb}
\usepackage[mathscr]{euscript} %To be able to use script letters with \mathscr{}
\usepackage{amsthm} %To be able to write Definitions, Theorems, etc.
\usepackage{tikz} %To draw graphs.
\usepackage{graphicx} % To scale equations
\newcommand*{\Scale}[2][4]{\scalebox{#1}{\ensuremath{#2}}} % To scale equations
\usepackage{framed} % To be able to frame theorems and stuff
%\usepackage{float} %To be able to place figures exactly where I want with [H]
\usepackage{multirow}
\usepackage{color}
\usepackage{subfigure}
\usepackage{longtable}
\usepackage{MnSymbol}
\usepackage{xcolor}  \definecolor{shadecolor}{rgb}{.95,.95,.95}  %To put a shaded region
\usepackage[numbers,square, comma, sort&compress]{natbib}	 %Make the references [1-8], rather than [1]-[8], as with \usepackage{cite}
\renewcommand*{\bibfont}{\footnotesize} %Size of the references; only needed with natbib
%\usepackage{makeidx} %To create an index
%\makeindex
\usepackage[hidelinks]{hyperref} %To be able to use links inside the document; [hidelinks removes the ugly red boxes]

%==================================================== ALGORITHM SETUP
\usepackage[ruled,vlined]{algorithm2e}

%========================================================================= THEOREMS, LEMMAS, DEFINITIONS, ETC.
\newtheorem{myDefinition}{Definition}
\newtheorem{myTheorem}{Theorem}
\newtheorem{myLemma}{Lemma}
\newtheorem{myCorollary}{Corollary}
\newtheorem{myProposition}{Proposition}
\newtheorem{myExample}{Example}
\newtheorem{myRemark}{Remark}
\newtheorem{myConjecture}{Conjecture}
\newtheorem{myQuestion}{Question}
\newtheorem{myProblem}{Problem}

%========================================================================= STUFF TO DRAW GRAPHS
\tikzstyle{every edge}=  [draw]
\tikzstyle{vertex} = [draw,circle,minimum size=1pt]
\tikzstyle{label} = [minimum size=.1pt,font=\scriptsize]
\tikzstyle{title} = [minimum size=.25cm,font=\small]

%========================================================================= SYMBOLS
%\newcommand{\var}[1]{\mathsf{#1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\wt}[1]{\widetilde{#1}}
\newcommand{\careful}{\textcolor{red}}
\newcommand{\comment}{\textcolor{blue}}

%Miscelaneous
%\def \R{\mathbb{R}}
%\def \P{\mathsf{P}}
%\def \T{\mathsf{T}}
%\def \c{\mathsf{c}}
\def \spn{{\rm span}}
\def \rnk{{\rm rank}}

%Scalars
\def \r{{\hyperref[rDef]{r}}}

%text
\def \as{{\hyperref[asDef]{{\rm a.s.}}}}
%Assumptions
\def \Cone{{\hyperref[ConeDef]{{\rm {\bf C1}}}}}
%Theorems
\def \mainThm{{\hyperref[mainThm]{Theorem \ref{mainThm}}}}
%Sections
\def \modelSec{{\hyperref[modelSec]{Section \ref{modelSec}}}}
%Figures
\def \distributionFig{{\hyperref[distributionFig]{Figure \ref{distributionFig}}}}
%Algorithms
\def \validationAlg{{\hyperref[validationAlg]{Algorithm \ref{validationAlg}}}}

\newcounter{@inst}
\newcounter{@auth}
\newcounter{auco}
\newdimen\instindent
\newbox\authrun
\newtoks\authorrunning
\newtoks\tocauthor
\newbox\titrun
\newtoks\titlerunning
\newtoks\toctitle

\def\clearheadinfo{\gdef\@author{No Author Given}%
                   \gdef\@title{No Title Given}%
                   \gdef\@subtitle{}%
                   \gdef\@institute{}%
                   \gdef\@thanks{}%
                   \global\titlerunning={}\global\authorrunning={}%
                   \global\toctitle={}\global\tocauthor={}}

\def\institute#1{\gdef\@institute{#1}}

\def\institutename
 {\begingroup
 \if!\@institute!\else
 \def\thanks##1{\unskip{}}%
 \def\iand{\\[5pt]\let\and=\nand}%
 \def\nand{\ifhmode\unskip\nobreak\fi\ $\cdot$ }%
 \let\and=\nand
 \def\at{\\\let\and=\iand}%
 \footnotetext[0]{\kern-\bibindent
 \ignorespaces\@institute}\vspace{5dd}\fi
 \endgroup
 }%

% correct bad hyphenation here
%\hyphenation{op-tical net-works semi-conduc-tor}

@@

\title{Prediction of functional markers of mass cytometry data via
  deep learning}
% Use \titlerunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
%\thanks{Both authors contributed equally to this manuscript}
\author{Claudia Sol\'is-Lemus, Xin Ma, Maxwell Hostetter II, Suprateek
  Kundu, Peng Qiu, Daniel Pimentel-Alarc\'on}
% Use \authorrunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\institute{Claudia Sol\'is-Lemus \at Emory University, Atlanta, GA, \email{csolisl@emory.edu}
\and Xin Ma \at Emory University, Atlanta, GA,
\email{xin.ma@emory.edu}
\and Maxwell Hostetter II \at Georgia State University, Atlanta, GA,
\email{mhostetter1@student.gsu.edu}
\and Suprateek Kundu \at Emory University, Atlanta, GA,
\email{suprateek.kundu@emory.edu}
\and Peng Qiu \at Georgia Institute of Technology and Emory University, Atlanta, GA,
\email{peng.qiu@bme.gatech.edu}
\and Daniel Pimentel-Alarc\'on \at Georgia State University, Atlanta, GA,
\email{pimentel@gsu.edu}}
%
% Use the package "url.sty" to avoid
% problems with special characters
% used in your e-mail or web address
%
\maketitle

\begin{abstract}
Recently, there has been an increasing interest in the analysis of
flow cytometry data, which involves measurements of a set of surface
and functional markers across hundreds and thousands of cells. These
measurements can often be used to differentiate various cell types and
there has been a rapid development of analytic approaches for
achieving this. However, in spite of the fact that measurements are
available on such a large number of cells, there have been very
limited advances in deep learning approaches for the analysis of flow
cytometry data. Some preliminary work has focused on using deep
learning techniques to classify cell types based on the cell protein
measurements. In a first of its' kind study, we propose a novel deep
learning architecture for predicting functional markers in the cells
given data on surface markers. Such an approach is expected to
automate the measurement of functional markers across cell samples,
provided data on the surface markers are available, that has important
practical advantages. We validate and compare our approach with
competing machine learning methods using a real flow cytometry
dataset, and showcase the improved prediction performance of the deep
learning architecture.
\end{abstract}

@@

Despite the technological advances in acquiring an increasing number
of parameters per single cell, approaches for analyzing such complex
data lag behind. The existing approaches are often subjective and
labor-intensive. For example, the widely used gating approach
identifies cell types by user-defined sequences of nested 2-D
plots. There have been efforts to develop clustering algorithms (e.g.,
flowMeans\cite{Aghaeepour2010}, flowSOM\cite{VanGassen2015},
X-shift\cite{Samusik2016}, and dimension reduction algorithms
(e.g., SPADE \cite{Qiu2011}, tSNE\cite{vanderMaaten2008},
Scaffold\cite{Spitzer2015}. However, there is still huge space
for developing new methods to ask new questions in this field.
% ---

@@

Furthermore, CyTOF data is perfectly suited for deep learning methods.
On one side, identify markers define a cell type (e.g., B cell, T
cell, monocytes, MSC), and on the other side, expressions of
functional markers identify the cell's activity (e.g., quiescent,
secreting cytokines, proliferating, apoptosis). Since CyTOF technology
allows for the simultaneous measurement of a large number of protein
markers, most CyTOF studies measure both identity markers and
functional markers, providing data for supervised learning tools, like
neural networks.  In addition, each CyTOF run typically collects data
on $10^6$ cells, creating an ideal large dataset in which the number
of samples (cells) is orders of magnitude larger than the number of
variables (markers). Deep learning methods are particularly suited for
this type of big data.

In terms of motivation, there are two main reasons to predict the functional markers from surface markers in CyTOF data: 1) monetary and time cost, and 2) technical limit of the total number of markers CyTOF can measure, which is currently around 50 protein markers. That is, if we can accurately predict some functional markers based on surface markers, there is no longer the need to include those functional markers in the staining panel (experimental design), and thereby freeing up channels to measure more surface markers or additional functional markers that cannot be predicted.

@@

\begin{figure*}
\centering
\includegraphics[scale=0.35]{figures/surface-hist.pdf}
\includegraphics[scale=0.35]{figures/functional-hist.pdf}
\includegraphics[scale=0.35]{figures/scatterplot.pdf}
\caption{Exploratory plots of surface and functional markers. The
  histograms show a biased pattern, and the scatterplot shows non
  linearity, both violations of crucial assumptions in standard
  regression models.}
\label{exploratory}
\end{figure*}

@@

The CyTOF dataset has been previously published in
\cite{Bendall2011,Qiu2011}. It contains single-cell data for 5 bone marrow samples from healthy donors. The data for each sample contains measurements for 31 protein markers for individual cells, including 13 cell surface markers which are conventionally used to define cell types, as well as 18 functional markers which reflect the signaling activities of cells. The number of cells per sample is roughly 250,000, and the total number of cells across all 5 samples is 1,223,228. Thus, the data can be expressed in a $1223228 \times 31$
matrix.

The data was transformed with inverse hyperbolic sine function
($arcsinh$ with co-factor of 5), which is the standard transformation
for CyTOF data \cite{Bendall2011}.

@@

We will compare the performance of different methods (explained next
section) to predict the functional markers with the surface markers.
The data is highly complex and correlated, violating some of the
fundamental assumptions of standard statistical approaches (like
regression). For example, the data is highly skewed and the pattern
between response and predictor is not linear (see figure
\ref{exploratory}).

@@

The complete data consisted of 1,223,228 cells in 18 functional
markers (responses) and 15 surface markers (predictors), which we
divided as follows: 750,000 rows as training set, 250,000 rows as
validation set, and 223,228 rows as testing set.

@@

The vector MSE is defined as
\begin{equation}
MSE_{vec} = \frac{1}{2n} \sum_{i=1}^n ||\hat{Y}_i - Y_i ||^2_2
\label{vectorMSEformula}
\end{equation}
where $\hat{Y}_i \in \mathbf{R}^{18}$ is the predicted vector of
responses for individual $i$, and $Y_i \in \mathbf{R}^{18}$ is the
observed vector of responses for individual $i$.

The individual MSE for predictor $k$ is defined  as
\begin{equation}
MSE_{(k)} = \frac{1}{2n} \sum_{i=1}^n (\hat{Y}_{k,i} - Y_{k,i} )^2
\label{MSEformula}
\end{equation}
where $\hat{Y}_{k,i} \in \mathbf{R}$ is the $k^{th}$ predicted
response ($k=1,\cdots,18$) for individual $i$, and $Y_{k,i} \in
\mathbf{R}$ is the $k^{th}$ observed response for individual $i$.

\begin{figure*}
\centering
\includegraphics[scale=0.4]{figures/neuralNet.eps}
\caption{Neural network for predicting functional markers (18
  responses) from surface markers (15 predictors) with 4 hidden
  layers with 90, 90, 45 and 45 nodes each.}
\label{nn}
\end{figure*}

% https://github.com/pluskid/Mocha.jl/blob/master/docs/user-guide/solver.rst
\vspace{0.25cm}
\noindent \textbf{Neural network model:} We tested different network
architectures, activation functions, regularization coefficients,
solver methods, momentum policies, and learning rates with 50,000 maximum epochs. The best
network has four layers (see figure \ref{nn}) with 90, 90, 45 and 45
nodes. The network uses hyperbolic tangent as activation function,
regularization coefficient of 0.0001, momentum policy fixed at 0.8, inverse-decay learning
rate policy with base learning rate, gamma and power parameters at 0.01, 0.0001, 0.75. We used Adam solver\cite{Kingma2014}, an
algorithm for first-order gradient-based optimization of stochastic
objective functions, based on adaptive estimates of lower-order
moments, instead of Stochastic Gradient Descent (SGD) as the former
showed increased accuracy. All networks were trained using the julia
package \texttt{Mocha}\cite{Bezanson2017,mocha}.

\vspace{0.25cm}
\noindent \textbf{Linear regression (unpenalized/penalized):} We fit
standard linear regression, as well as the penalized version with
LASSO penalty under different penalization parameters. We used the
\texttt{ScikitLearn}\cite{scikit} julia wrapper, with default
settings. We noted that the penalized version performed worse than the
unpenalized version for all the predictors (regardless of penalty
parameter), so we only present results below for the unpenalized
linear regression model.

\vspace{0.25cm}
\noindent \textbf{Decision tree and random forest regressions:} We fit one decision tree
regression per response with \texttt{ScikitLearn} julia wrapper, with default
settings. We compared the performance of the ``mse'' criterion and the
Friedman's improvement score, deciding on the former (``mse'') which
is the default setting. We did not constraint the maximum depth of the
tree, and set as 2 the minimum number of samples required to split an
internal node. In addition, we did not constraint the maximum number
of features to consider when looking for the best split. Later, we fit
20 trees into a random forest regression. We could not explore more than 20
trees due to computational time constraints.

@@

\begin{figure}
\centering
\includegraphics[scale=0.45]{figures/vector-mse.pdf}
\includegraphics[scale=0.45]{figures/time.pdf}
\caption{Left: Vector MSE (equation \ref{vectorMSEformula}) for all
  four methods sorted from most accurate (neural network) to least
  accurate (decision tree). Right: Running time (in seconds) for the
  training and validation sets (sample 750,000 rows) for all four
  methods, sorted from fastest (linear regression) to slowest (random
  forest).}
\label{vectorMSE}
\end{figure}

Figure \ref{MSE} shows the individual MSE (equation \ref{MSEformula})
per response (18 responses in x-axis) for each of the four
methods. Again, the prediction accuracy of neural network is better
than the other three methods for all the 18 predictors.

\begin{figure}
\centering
\includegraphics[scale=0.45]{figures/mse.pdf}
\caption{Individual MSE (equation \ref{MSEformula}) for each 18
  responses. The neural network outperforms all other methods across
  all responses.  Lines are drawn simply for visual effect.}
\label{MSE}
\end{figure}

The MSE performance varies across responses. For example, the first response (functional marker 141.pPLCgamma2) has an overall MSE lower than other responses like the third (functional marker 152.Ki67), the 8th (functional marker 159.pSTAT3) or the 14th (functional marker 171.pBtk.Itk).

Figure \ref{violin} (left) shows the violin plots for these 4 functional markers. We observe that the 14th response has a wider range and heavier tails than the other responses, which is confirmed in the scatterplots on the center and right (figure \ref{violin}). It appears that the wider spread and higher variability of the 14th response (functional marker 171.pBtk.Itk) causes the lower prediction accuracy compared to other responses, like the first one (functional marker 141.pPLCgamma2).

\begin{figure*}
\centering
\includegraphics[scale=0.35]{figures/violin.pdf} \hspace{0.25cm}
\includegraphics[scale=0.35]{figures/response1.pdf}
\hspace{0.25cm}
\includegraphics[scale=0.35]{figures/response14.pdf}
\caption{Left: Violin plot for four functional markers (responses). Horizontal lines represent the 25th quantile, median and 75th quantile. Center: Predicted vs observed responses on the first functional marker (141.pPLCgamma2)
  across all four methods. Left: Predicted vs observed responses on the 14th functional marker (171.pBtk.Itk)
  across all four methods. The closer the slope to 1 (black line), the
  better.}
\label{violin}
\end{figure*}

Finally, we present selected scatterplots of surface markers as predictors for the responses 1,3,8 and 14 (figure
\ref{predObs}). We can appreciate in these plots the non-linear relationship between the predictors and responses, which justifies the use of a neural network approach.

\begin{figure}[ht]
\centering
\includegraphics[scale=0.3]{figures/scatterplot1.pdf} \hspace{0.5cm}
\includegraphics[scale=0.3]{figures/scatterplot14.pdf}
\caption{Scatterplot of selected surface markers (predictors) and selected functional markers (responses). Left: the first response (functional marker 141.pPLCgamma2) shows a linear relationship to the predictor (surface marker 115.CD45), which partially explains the better MSE in figure \ref{MSE}. Right: the 14th response (functional marker 171.pBtk.Itk) shows a non-linear relationship to the predictor (surface marker 115.CD45), which partially explains the worse MSE in figure \ref{MSE}.}
\label{predObs}
\end{figure}

@@

The improved prediction accuracy of neural networks can be explained
by their flexibility to account for non-linearity or skewness.  Unlike
regression models, neural networks do not have linearity or normality
assumptions, and they take advantage of the correlation structure
among responses by fitting a network for the whole response vector.

As mentioned already, CyTOF data is perfectly suited for deep learning
methods given the simultaneous measurement of a large number of
protein markers, including both identity markers and functional
markers. Both measurements allow for the implementation of highly
accurate supervised methods, like neural networks. In addition, the
structure of CyTOF data is ideal for deep learning: number of samples
orders of magnitude greater than the number of variables.

The accuracy in the prediction of functional markers from surface markers has economic and computational advantages, for example, considering the limitation to the total number of markers CyTOF can measure, which is currently around 50 protein markers. Being able to predict functional markers from surface markers could allow for different types of staining panels which could measure more surface markers, or focus on functional markers not so easily predicted.

@@

% %======================================================= MAIN THEOREM
% \begin{framed}
% \begin{myTheorem}
% \label{mainThm}
% Write Theorem
% \end{myTheorem}
% \end{framed}
